Type Error Telemetry at Scale

### 2023-09-30 programming23/paper.tex
sec 4

% TODO BEFORE SUBMISSION
% be consistent between "IDs" and "ids" [SK prefers "IDs" to avoid it being read as a word]
% even more so between "ID" and "id" [NB: "id" is a word in English]
% \usepackage[sort]{natbib} or otherwise manually fix cites to be in ascending order

@@

This telemetry design is admittedly coarse-grained.
For instance, it does not distinguish between edits that ignore a type
error from edits that remove one error while introducing another.
We acknowledge this and other threats in~\cref{s:threats}.
The main advantages of this telemetry are the small size of each record
and complete the lack of private information.
Furthermore, despite the limitations, this telemetry supports a variety of
interesting conclusions that we showcase in the next section.

@@

Roblox Studio collected type error telemetry between February
and April 2023.
Every Studio session had a small random chance of generating
telemetry~(\cref{s:telemetry-records}).
The chosen sessions generated a record whenever the user switched modules and
randomly on each keystroke.
In total, we collected over 1.5 million telemetry records.
Roughly two thirds of records are due to keystrokes.

\Cref{f:records-per-hour} provides a time-ordered distribution of the data.
Each vertical bar counts the number of records generated per hour according to
a client-side timestamp.
%% \todo{Do we need all this detail?}
%% the vertical bars suggest we have lots of day-job developers, will point this out below
The bars are labeled with a California time zone because that is the locale of
the Roblox servers, which parsed the timestamps.
Most buckets contain roughly one thousand records, but the number
varies between a few dozen at minimum at over three thousand.
%% \FILL{} investigate one tower, what's happening?? all one session?
There is a sharp drop in early April when the experiment was finished
and the telemetry probe was removed, though there is a long tail as
developers updated {Roblox Studio}.
We do not know which timezone the records originated in, but since the counts
tend to peak midday California time it seems that the majority
of Roblox developers are following a Western Hemisphere schedule.
The regular pattern of the peaks furthermore suggests that many creators
follow a regular schedule; this may be due to a large community of
professional developers.
The counts for weekends (shaded regions) are often the tallest,
as Roblox has a significant school-aged creator community.

@@

A close inspection of the data revealed two anomalies:
\begin{enumerate}
  \item
    Some records have identical timestamps and session IDs.
    We keep the first such record and discard the rest.
    (A human would have to switch modules twice or enter several keystrokes
     within one millisecond to generate two such records legitimately.)

@@

Three important characteristics of the recorded data are the size of the codebases,
the length of the sessions, and the number of type
errors.
We discuss these in turn.
Codebase size and session length have long-tailed distributions
with many small items and few very large ones.
%% TODO check %% follow power-law distributions~\cite{csn-siamr-2009} 
%% https://aaronclauset.github.io/powerlaws/
%% bin version, is this relevant? sounds like no from abtract https://arxiv.org/pdf/1208.3524.pdf
The number of type analysis errors is somewhat low~(explained in the next
section), but sufficient to proceed with futher analysis.

@@

\Cref{t:codebase-size} summarizes the size of projects in the dataset.
The middle four columns report the mean, standard deviation, median,
and \percentile{99} for the number of lines
and lines in edit range.\footnote{The data also contains the number of files in
each project, but these numbers are difficult to relate to typechecking because
projects may contain hundreds of files that define data assets.  For the
record, the median file count is 7,678 and the \percentile{99} is 51,761 files.
}
The plots on the right show zoomed-in distributions of the line and edit range
sizes.
For example, the x-axis of the first plot counts up to 1,000 lines (not showing
a long tail to the right) and the y-axis counts up to 1,000 telemetry records
(not showing a tall spike on the left).

@@

\Cref{t:session-size} outlines the size of sessions using two metrics:
real time in seconds between the first and last record, and the total number of
records.
The middle columns report the mean, standard deviation,
median, and \percentile{99}.
Similar to the previous table~(\cref{t:codebase-size}), the distributions focus
on the lower-left fragment of the data, which again has a very long tail.
%% comment on long-tail record counts?
The longest sessions last several days and/or contain thousands of records,
while the shortest last a few milliseconds or include just one record.
% Likewise, both plots in the figure follow a power-law distribution
% (especially for record count).
A median session runs on the order of minutes and consists of a few dozen events.
%% The median time is 14 minutes and the median 
%% The median number of records is 138.
% These numbers are reasonable.

@@

The large fraction of errors in the current module is encouraging
for two reasons.
First, it suggests that errors appear locally, as the result of edits to nearby
code, rather than as the result of edits to far-away code that lives in another
module.
Second, it suggests that creators typically fix errors before switching
to another module.

@@

  \item
    On the other hand, the low overlap between errors (whether current or previous)
    and edits says that edits rarely target highlighted code.
    There are several possible explanations:
    analysis might report several errors and the edits only target one or two of them;
    creators might ignore errors entirely and edit other code; or
    creators might choose to edit a non-highlighted location to fix an error.
    Our data cannot narrow down the answer.

@@

\Cref{f:error-by-mode} divides errors across modes.
By the left plot, \mnonstrict{} accounts for two thirds of all type errors.
This confirms our expectation.
%% because \mnocheck{} reports only syntax errors and
%% \mstrict{} covers only a small number of telemetry records.
The right plot, by contrast, does not support our hypothesis that \mnocheck{}
records have more \FS{} errors than the other modes.
Though \mnocheck{} does have the most \FS{} errors, its proportion matches
the proportion of \mnocheck{} records in the
dataset~(\cref{f:total-records})---if every record had exactly one \FS{} error,
the plot would be the same.

Digging further into \FS{} errors, the median number of such errors by mode is
1 in \mnocheck{}, 3 in \mnonstrict{}, and 1 in \mstrict{}.
Records in \mnocheck{} mode evidently do not have significantly more \FS{}
errors.
Furthermore, type analysis does not seem to reduce \FS{} errors.

@@

The main focus of our telemetry design is to learn how type errors intersect
with edit ranges.
Indeed, most of the fields in a telemetry record are dedicated to this
topic~(\cref{f:tdata}), and it helps to answer RQ2 from the introduction:
\begin{description}
  \item[RQ2.]
    For modules that use type analysis:
    which errors arise,
    how do sessions respond,
    and which errors tend to persist despite subsequent edits?
\end{description}

@@

All three counts are imprecise because our telemetry system cannot distinguish
particular type errors.
If a creator removes one \code{NotATable} error but then
immediately introduces another on a different line,
then the centor (\keepsym{}) column gets incremented
instead of the increase (\addsym{}) and decrease (\dropsym{}) columns.

@@

  \item
    \code{SyntaxError} and the related errors \code{UnknownSymbol} and
    \code{UnknownProperty} are the most popular.
    If the edit range errors are indicative of type errors at large,
    then most errors in the data are likely due to typos.
    These errors also survive (\keepsym{}) the most often.

    \subitem The random-keystroke strategy of our telemetry may be inflating
    the number of typos because it has no way to avoid reporting edits that
    happen in the middle of things.
    For example, a creator who writes a method call
    \code{bucket.countFish()} letter-by-letter will see errors
    including \code{UnknownSymbol(buck)} and \code{UnknownProperty(bucket.c)}.
    Telemetry may pick these up;~\cref{s:conclusion} suggests ways
    to avoid doing so in the future.

  \item
    After the syntax errors, \code{CountMismatch}, \code{TypeMismatch}, and
    \code{GenericError} are the next most common \mnonstrict{} and \mstrict{}
    errors.
    A \code{CountMismatch} sends the wrong number of arguments to a function.
    It may simply be a common error, but it may also be due to our random-keystroke
    telemetry.
    The high incidence of \code{TypeMismatch} and \code{GenericError} calls for
    further study in future work to determine precisely which error arose.
    (Some of the \code{GenericError}s come with context strings in the form of
    \code{ExtraInformation} errors, but telemetry does not report these strings.)
    %% GenericError = only a few things: extend non-class, iterate non-table, ...?

@@

  \item
    %% Drop >= Add >= Rem ~= 18
    %% other nonzero      ~= 21 triples
    Ideally, the highest number in each column group should
    be the number of decreases (\dropsym{}).
    A decrease should happen when the creator sees an error,
    edits around the highlighted code, and fixes it.
    However, decreases are highest in only 18 of the 39 nonzero groups.

@@

\Cref{t:type-error-count} reports the percent of type errors that overlap with
the edit range.
The purpose of this table is to discover all specific errors in the dataset,
thus it includes both keystroke-based and module-switch-based records---even
though edit ranges are not well-defined for the latter.

@@

\begin{itemize}
  \item
    Syntax errors, in particular \code{UnknownSymbol}, are much more
    common in \mnonstrict{} than \mstrict{}.
    This may be because \mnonstrict{} reports fewer errors in general.

@@

Fortunately, the data rarely contains too-complex errors.
None of these errors appear in edit ranges~(\cref{t:type-error-count}).
For the specific case of \code{CodeTooComplex}, our telemetry tracked project-wide
counts and found only 26 errors, which were spread across eleven
records in three sessions.
It is entirely possible that they all came from one session that
deliberately pushed the limits of type analysis.
%% Whatever the reason, the overall number is low enough to suggest that the
%% internal limits are not posing barriers to typechecker adoption.

@@

\begin{itemize}
  \item
    The type error increases and decreases are remarkably balanced: across all
    three modes, they are at most \pct{2} apart.
    While some of this balance is due to module-switch records
    (if the two modules have different error counts, there will be a change),
    most of the data is from keystrokes; thus it appears that sessions tend to
    fix errors highlighted by type analysis.

@@

\Cref{f:tefs-compass} takes a closer look at type errors (\tekey{}) and \FS{} errors (\fskey{}).
For every pair of events in each session, we compare the change in
type errors to the change in \FS{} errors and choose one of five categories.
At a high level, the question at hand is whether the changes \emph{agree} or \emph{disagree}.
The categories go one step further and separate zero and non-zero changes:

@@

>>>>

% TODO BEFORE SUBMISSION
% be consistent between "IDs" and "ids" [SK prefers "IDs" to avoid it being read as a word]
% even more so between "ID" and "id" [NB: "id" is a word in English]
% \usepackage[sort]{natbib} or otherwise manually fix cites to be in ascending order
% redo fig 5 to have better resolution

@@

This telemetry design is admittedly coarse-grained.
For instance, it does not distinguish between edits that ignore a type
error from edits that remove one error while introducing another.
We acknowledge this and other threats in~\cref{s:threats}.\todo{Do we?
Double check.}
The main advantages of this telemetry are the small size of each record
and complete the lack of private information.
Furthermore, despite the limitations, this telemetry supports a variety of
interesting conclusions that we showcase in the next section.

@@

We collected type error telemetry in Roblox Studio between February
and April 2023.
Every Studio session had a small random chance of generating
telemetry~(\cref{s:telemetry-records}).
The chosen sessions generated a record whenever the user switched modules and
randomly on each keystroke.
In total, we collected over 1.5 million telemetry records.
Roughly two thirds are due to keystrokes.

\Cref{f:records-per-hour} provides a time-ordered distribution of the data.
Each vertical bar counts the number of records generated per hour according to
a client-side timestamp.\todo{Not just per day?!?}
The bars are labeled with a California time zone.
Buckets\todo{What is a BUCKET as opposed to a BAR?}
contain a few dozen to over 3,000 records, typically about 1,000.
%% \FILL{} investigate one tower, what's happening?? all one session?
Telemetry ended in early April; the long tail is due to developers
updating Roblox Studio later.
% BEN: I see no basis for this; nor is this at all interesting.
% We do not know which timezone the records originated in, but since the counts
% tend to peak midday California time it seems that the majority
% of Roblox developers are following a Western Hemisphere schedule.
The regular pattern of the peaks suggests that many creators
follow a regular schedule; these may be
professional developers.
The tall weekend (shaded) regions could be because
Roblox has a significant school-aged creator community.

@@

A close inspection of the data revealed two anomalies:
\begin{enumerate}
  \item
    Some records have identical timestamps and session IDs.
    We keep the first such record and discard the rest.
    (A human would have to switch modules twice or enter several keystrokes
     within 1ms to generate two such records legitimately.)

@@

Three important characteristics of the recorded data are the size of the codebases,
the length of the sessions, and the number of type
errors.
We discuss these in turn.
% Commenting these out because that's literally the rest of the
% subsection. Say anything you want from here over there!
% Codebase size and session length have long-tailed
% distributions\todo{DID YOU DO A STATISTICAL TEST.}
% with many small items and few very large ones.
% %% TODO check %% follow power-law distributions~\cite{csn-siamr-2009} 
% %% https://aaronclauset.github.io/powerlaws/
% %% bin version, is this relevant? sounds like no from abtract https://arxiv.org/pdf/1208.3524.pdf
% The number of type analysis errors is somewhat low~(explained in the next
% section), but sufficient to proceed with futher analysis.

@@

\Cref{t:codebase-size} summarizes the size of projects in the dataset.
The four numeric columns report the mean, standard deviation, median,
and \percentile{99} for the number of lines
and lines in edit range.\footnote{The data also contains the number of files in
each project, but these numbers are difficult to relate to typechecking because
projects may contain hundreds of files that define data assets.  For the
record, the median file count is 7,678 and the \percentile{99} is 51,761 files.
}
The plots on the right show zoomed-in distributions of the line and edit range
sizes.
For example, the x-axis of the first plot counts up to 1,000 lines (not showing
a long tail to the right) and the y-axis counts up to 1,000 telemetry records
(not showing a tall spike on the left).

@@

\Cref{t:session-size} similarly outlines the size of sessions using two metrics:
real time in seconds between the first and last record, and the total number of
records.
The plots focus
on the lower-left fragment of the data, which again has a very long tail.
%% comment on long-tail record counts?
The longest sessions last several days and/or contain thousands of records,
while the shortest last a few milliseconds or include just one record.
% Likewise, both plots in the figure follow a power-law distribution
% (especially for record count).
A median session runs on the order of minutes and consists of a few dozen events.
%% The median time is 14 minutes and the median 
%% The median number of records is 138.
% These numbers are reasonable.

@@

The large fraction of errors in the current module is encouraging
for two reasons.
First, it suggests that errors appear locally, as the result of edits to nearby
code, rather than as the result of edits to code that lives in another
module.
Second, it suggests that creators typically fix errors before switching
to another module.

@@

  \item
    On the other hand, the low overlap between errors (whether current or previous)
    and edits says that edits rarely target highlighted code.
    There are several possible explanations:
    analysis might report several errors and the edits only target one or two of them;
    creators might ignore errors entirely and edit other code; or
    creators might choose to edit a non-highlighted location to fix an error.
    While non-privacy-protecting analysis have answered such
    questions~\cite{mfk-sigcse-2011}, our data cannot narrow down the answer.

@@

\Cref{f:error-by-mode} divides errors across modes.
By the left plot, \mnonstrict{} accounts for two thirds of all type errors.
This confirms our expectation.
%% because \mnocheck{} reports only syntax errors and
%% \mstrict{} covers only a small number of telemetry records.
The right plot, by contrast, does not support our hypothesis that \mnocheck{}
records have more \FS{} errors than the other modes.
Though \mnocheck{} does have the most \FS{} errors, its \emph{proportion} matches
the proportion of \mnocheck{} records in the
dataset~(\cref{f:total-records})---if every record had exactly one \FS{} error,
the plot would be the same.

Digging further into \FS{} errors, the median number of such errors by mode is
1 in \mnocheck{}, 3 in \mnonstrict{}, and 1 in \mstrict{}.
Records in \mnocheck{} mode evidently do not have significantly more \FS{}
errors.
Furthermore, type analysis does not seem to reduce \FS{}
errors.\todo{Isn't this a bit surprising?}

@@

The main focus of our telemetry design is to learn how type errors intersect
with edit ranges.
Indeed, most of the fields in a telemetry record are dedicated to this
topic~(\cref{f:tdata}), and are directed to RQ2 on type errors
(\cref{s:introduction}).

@@

% BEN: double-check, I think this says all you were trying to say.
% "typically" because we COULD have gotten lucky.
The counts are imprecise because of telemetry granularity.
If a creator removes one error but then
immediately introduces another on a different line, this typically manifests as
a \keepsym{} rather than one \addsym{} and one \dropsym{}.

@@

  \item
    \code{SyntaxError} and the related \code{UnknownSymbol} and
    \code{UnknownProperty} are most popular.
    If the edit range errors are indicative of type errors at large,
    then most errors in the data are likely due to typos.
    These errors also survive (\keepsym{}) the most often.

    Our telemetry strategy likely inflates
    the number of typos because it cannot determine when a creator is
    mid-edit.
    For example, a creator who writes a method call
    \code{bucket.countFish()} letter-by-letter will generate an
    \code{UnknownSymbol} error for \code{buck} and
    \code{UnknownProperty} for \code{bucket.c}, which may end up sampled.
    \Cref{s:conclusion} suggests ways
    to avoid this in the future.

  \item
    After the syntax errors, \code{CountMismatch}, \code{TypeMismatch}, and
    \code{GenericError} are the next most common \mnonstrict{} and \mstrict{}
    errors.
    A \code{CountMismatch} sends the wrong number of arguments to a function.
    It may simply be a common error, but it may also be a
    telemetry artifact.
    The high incidence of \code{TypeMismatch} and \code{GenericError} calls for
    further study in future work to determine precisely which error arose.
    (Some of the \code{GenericError}s come with context strings in the form of
    \code{ExtraInformation} errors, but telemetry does not report these strings.)
    %% GenericError = only a few things: extend non-class, iterate non-table, ...?

@@

  \item
    %% Drop >= Add >= Rem ~= 18
    %% other nonzero      ~= 21 triples
    Ideally, the highest number in each column group should
    be decreases (\dropsym{}).
    A decrease should happen when the creator sees an error and
    fixes it.
    However, decreases are highest in only 18 of the 39 nonzero groups.

@@

\Cref{t:type-error-count} reports the percent of type errors that overlap with
the edit range.\todo{I DO NOT UNDERSTAND ANYTHING HERE.}
The purpose of this table is to discover all specific errors in the dataset,
thus it includes both keystroke-based and module-switch-based records---even
though edit ranges are not well-defined for the latter.

@@

\begin{itemize}
  \item
    Syntax errors, in particular \code{UnknownSymbol}, are much more
    common in \mnonstrict{} than \mstrict{}.\todo{This is NOT a
      ``syntax'' error? It's a typo!}
    This may be because \mnonstrict{} reports fewer errors in general.

@@

Fortunately, the data rarely contains too-complex errors.
None of these errors appear in edit ranges~(\cref{t:type-error-count}).
For the specific case of \code{CodeTooComplex}, our telemetry tracked project-wide
counts and found only 26 errors, which were spread across eleven
records in three sessions.
Note that all three sessions \emph{could} have been from the same codebase.
% BEN: I see no reason for this speculation. The only important thing
% is the fact that this could have been all the same codebase.
%It is entirely possible that they all came from one session that
%deliberately pushed the limits of type analysis.
%% Whatever the reason, the overall number is low enough to suggest that the
%% internal limits are not posing barriers to typechecker adoption.

@@

\begin{itemize}
  \item
    The type error increases and decreases are remarkably balanced: across all
    three modes, they are at most \pct{2} apart.
    While some of this balance is due to module-switch records
    (if the two modules have different error counts, there will be a change),
    most of the data is from keystrokes; thus, it appears that sessions tend to
    fix errors highlighted by type analysis.

@@

\Cref{f:tefs-compass} takes a closer look at type errors (\tekey{}) and \FS{} errors (\fskey{}).
\todo{I'm kinda' baffled.}
For every pair of events in each session, we compare the change in
type errors to the change in \FS{} errors and choose one of five categories.
At a high level, the question at hand is whether the changes \emph{agree} or \emph{disagree}.
The categories go one step further and separate zero and non-zero changes:

@@

\newpage
